---
title: "Plotting a Network Graph using network_p2p"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: "lumen"
---

This script demonstrates how to generate and visualize a network graph using the `network_p2p` function. The function creates an `igraph` object, which can be plotted to display connections between individuals based on collaboration metrics.

## Step 1: Load libraries and sample data

In this example, we will use the sample `p2p_data_sim()` dataset from the **vivainsights** package. We will also use **dplyr** for data manipulation, **igraph** for network graph creation, **ggplot2** for visualization, and **ggraph** for enhanced network plotting.

```{r setup, message=FALSE, warning=FALSE}
library(igraph)
library(ggplot2)
library(RColorBrewer)
library(ggraph)
library(dplyr)
library(vivainsights)

# Define HR variable
hrvar_text <- "Organization"

# Display the first few rows of the dataset
head(p2p_data_sim())
```

## Step 2: Generate the igraph network object

The `network_p2p()` function constructs a network graph based on collaboration data. We set:

-   `data` to the simulated P2P dataset
-   `hrvar` to define the grouping attribute
-   `return = "network"` to get an igraph object

```{r}
g <- network_p2p(
  data = p2p_data_sim(), 
  hrvar = hrvar_text, 
  return = "network")

# Ensure g is an igraph object
if (!inherits(g, "igraph")) {
  stop("network_p2p did not return an igraph object. Check function parameters.")
}
```

## Step 3: Prepare and customize the graph for visualization

Before plotting, we refine the graph by:

-   Removing `loops` (self-connections) and `multiple edges` (redundant links)
-   Extracting unique values for color mapping
-   Assigning colors and scaling node sizes

```{r}
# Simplify the graph (remove redundant edges and self-loops)
g <- simplify(g, remove.multiple = TRUE, remove.loops = TRUE)

# Extract unique values for color mapping
unique_values <- unique(V(g)$Organization)
num_unique_values <- length(unique_values)

# Generate a color palette
colors <- brewer.pal(min(num_unique_values, 8), "Set2")
org_to_color <- setNames(colors, unique_values)

# Assign colors and scale node sizes
V(g)$node_color <- org_to_color[V(g)$Organization]
V(g)$node_size <- V(g)$node_size * 120  # Ensure this attribute exists
```

## Step 4: Customize and plot the network graph

We use `ggraph` to create a visually appealing graph with:

-   `Edge color` → blue
-   `Vertex color` → mapped to organization groups
-   `Vertex size` → scaled according to `node_size`
-   `Theme` adjustments for a dark background and enhanced readability

```{r}
ggraph(g, layout = "mds") +
  geom_edge_link(aes(edge_alpha = 0.3), color = "#1f78b4") +
  geom_node_point(aes(size = V(g)$node_size, color = factor(V(g)$Organization))) +  
  scale_color_manual(values = org_to_color) +  
  theme_minimal(base_family = "sans") +  # Use a generic, available font
  theme(
    plot.background = element_rect(fill = "white", color = NA),
    panel.grid = element_blank(),  
    legend.text = element_text(size = 10, color = "black"),
    legend.background = element_rect(fill = "white"),
    legend.key = element_rect(fill = "white")
  ) +
  guides(color = guide_legend(override.aes = list(size = 5)), size = FALSE)
```

This final plot displays a network of peer-to-peer collaborations based on organization groupings, using a structured and aesthetically refined visualization.