---
title: "Introduction to Organizational Network Analysis with vivainsights in R"
author: ["Julia Kasper", "Martin Chan"]
date: "2023-11-02"
categories: ["R"]
tags: ["network analysis", "ONA"]
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
   collapse = TRUE,
   comment = "#>",
   fig.width=8,
   fig.height=8
  )
```

```{r message=FALSE, warning=FALSE, include=FALSE}
library(vivainsights)
library(dplyr)
```

This article provides a beginner's introduction to Organizational Network Analysis (ONA), its basic concepts, and how to run them with the functions available from the [**vivainsights** R library](https://microsoft.github.io/vivainsights/). We will address the following questions in this article:

- What is Organizational Network Analysis (ONA)?
- What are the basic building blocks of networks?
- What are the scenarios where ONA can be applied?
- What are the functions and outputs available in the **vivainsights** R library?

This is an updated version of the [original article](https://microsoft.github.io/wpa/articles/network-analysis.html) published for the legacy **wpa** library in 2021 August.

## Introduction

**Organizational Network Analysis (ONA)** is a family of methods to understand patterns of collaboration by examining the strength, frequency and nature of interactions between people. Analyzing and visualizing network connectivity in your organization can help you shape business strategy that unlocks success for individual and team productivity, innovation, employee engagement and organizational change, thereby helping your business become more resilient and effective.

Most valuable information in organizations is not codified. As employees collaborate, information flows from one person to the other or they generate knowledge together. These connections deliver value when information is exchanged and visualizing these relationships can provide us with a holistic view of how information flows through an organization. ONA can help us reveal central network collaborators, critical connections, and potential barriers to information flow.

### Basic Building Blocks

We will start with the basic building blocks of networks. Networks consist of two primary aspects, a multitude of separate entities ("nodes") and the connections between them ("edges"). Nodes and edges provide a foundation for understanding how information flows in your organization, and also strategizing on how they can flow and should flow. Nodes in ONA visualizations typically represent individual employees or groups of employees (e.g. `Organization`, `Region` etc.), serving as important hubs for the exchange of ideas and information. In addition to these nodes, you have connections between them called **edges**. These edges represent a connection between nodes, which in turn represents a collaboration relationship between individuals or groups. In the context of Viva Insights, these connections between nodes in the network represent collaboration activities (emails, meetings, calls, instant messages). In a network analysis, we also need to understand the strength of the tie. For this calculation we are using time spent together in collaboration. In addition, we have to consider the **direction** in which collaboration is operated. For meetings there is no direction, but for emails, chats and calls we have a sense of direction which we can look at.

> Viva Insights provides a powerful data source for performing network analysis and for surfacing insights about information flows in your organization. The **vivainsights** R library contains several powerful features that enables you to create such analysis and visualizations. 

## Scenarios

Network analysis provides a data-driven approach to address organizational challenges. Practical scenarios include:

- **Identifying silos**: ONA can help reveal critical silos where collaboration improvements can be generated. By tapping into this data, organizations can rebalance collaboration loads, expand capacity, and integrate expertise on the edge of the network to break down those silos. 
- **Identifying influencers**: By understanding the depth and intensity of collaboration, organizations can identify influencers. To boost employee engagement and prevent potential attrition, they can reduce collaborative overload on those central collaborators in the network and better leverage top talent.
- **Workspace planning**: ONA can surface the 'organic' patterns of how individuals and groups collaborate with each other, and help decision-makers determine how workspaces and seating should be allocated as organizations move back to the office.
- **Improve organizational culture**: Through insights into where information flows efficiently and flags where it flows inefficiently, this approach offers a great opportunity to reduce effort and improve value creation. With this information, key work activities can be enhanced, innovation can be driven, and network behaviour can be aligned to organizational strategies.

## Functions and Metrics

Network analysis is quantitative and uses collaboration data to create a matrix of relationships and applies matrix algebra to calculate measures. The **vivainsights** R package provides a set of functions that are designed to run these analyses and visualizations with the data from Viva Insights. Specifically, the data refers to two cross-collaboration queries, the [Group-to-group query](https://learn.microsoft.com/en-us/viva/insights/advanced/analyst/cross-collaboration-g2g) and the [Person-to-person query](https://learn.microsoft.com/en-us/viva/insights/advanced/analyst/cross-collaboration-p2p). 

1. **Group-to-group query**: captures collaboration between two groups in your organisation. For example, how often employees in design collaborate with US-based employees.
2. **Person-to-person query**: captures collaboration between individuals in your organisation. For example, how strongly managers in engineering are connected to managers in design. 

The corresponding functions from the package are as follows: 

1. [`network_g2g()`](https://microsoft.github.io/vivainsights/reference/network_g2g.html): an advanced function which manipulates a group-to-group collaboration graph. It creates a network plot with the group-to-group query and displays collaboration across your organization. The default metric value is `Group_collaboration_time_invested`, but this can be customized by supplying the name of the metric.

2. [`network_p2p()`](https://microsoft.github.io/vivainsights/reference/network_p2p.html): this function analyses a person-to-person (P2P) network query, with multiple visualisation and analysis output options (e.g. sankey plot, network plot). This function enables you to perform community detection as well as compute centrality statistics on the P2P network. 
   > Strong and diverse tie metrics are based on collaboration activities and indicates how varied, broad and strong a person’s connections or engagements are. See more from [here](https://learn.microsoft.com/en-us/viva/insights/advanced/reference/metrics).

## Queries and Outputs

The functions from the **vivainsights** library operate on top of queries generated from the Viva Insights Analyst experience. These queries are first generated, exported as CSV, and loaded into R as `data.frame` objects. These data frames are then passed into the functions as input parameters.

Another parameter that can be specified is the `return` parameter, which for instance controls whether the function would return a table (`return = "table"`) or a plot (`return = "plot"`). 

### Group-to-group 

Here is an example of the `network_g2g()` function run on the `g2g_data` data frame, with the `return` parameter set to `"plot"`.

```{r message=FALSE, warning=FALSE}
network_g2g(data = g2g_data, return = "plot")
```

- You can specify the metric being used for this visualization (Default: metric = "Group_collaboration_time_invested")
- Optionally you can provide a data frame that displays the size of each organization in the collaborator attribute. More information find [here](https://microsoft.github.io/vivainsights/reference/network_g2g.html).

*Interpretation: This visualization shows how strongly different organizations are connected. Each node represents a group in an organization, and the width of the edges represent the proportion of collaboration that occurs between the two groups. Note that an exclusion threshold is applied here, so that only collaboration proportions (edge widths) above a specified percentage are shown. Groups or nodes that do not have any edges connected to them ('islands') may have a possible risk of being siloed because they are not as strongly connected to other groups.* 

To display all the edges without setting an exclusion threshold, you can run: 
```{r message=FALSE, warning=FALSE}
network_g2g(data = g2g_data, exc_threshold = 0, return = "plot")
```

To better understand the data, you can also run the following code to return a table of the data: 

To validate hypotheses and better understand the data, it is also recommended that you run the following code to examine the interaction matrix. The interaction matrix is a matrix that captures the aggregated collaboration between all groups against each other, in the scope of the generated query. 

```{r message=FALSE, warning=FALSE} 
network_g2g(data = g2g_data, exc_threshold = 0, return = "table")
```

### Person-to-person

Running the network visualization for the Person-to-person query is similar. In the code below, we start off by generating a person-to-person data frame using the `p2p_data_sim()` function. This function generates a simulated data frame that is compatible with the `network_p2p()` function. Next, we run the `network_p2p()` function and specify the `return` parameter to be `"plot"`.

```{r message=FALSE, warning=FALSE}
p2p_data <- p2p_data_sim(size = 50)

p2p_data %>% network_p2p(hrvar = 'Organization', return = "plot")
```

Each node represents an individual, and the colour of the nodes correspond to the HR attribute specified in the arguments ('Organization'). An edge represents there being a reciprocal interaction between two individuals. By default, multi-dimensionality scaling (`layout = 'mds'`) is used for determining the layout of the network. 

### Centrality

One additional functionality of `network_p2p()` is the ability to generate centrality statistics. Centrality statistics are a set of metrics that are used to identify the most important nodes in a network. The following code generates a centrality table, which is a table of centrality statistics for each node in the network. 

```{r message=FALSE, warning=FALSE}
tb_centrality <- p2p_data %>%
   network_p2p(
      hrvar = 'Organization',
      centrality = 'degree',
      return = "data"
      )

head(tb_centrality)
```

It is also possible to run the above, averaged by an attribute specified with the `hrvar` argument: 

```{r message=FALSE, warning=FALSE}
p2p_data %>% network_p2p(hrvar = 'Organization', centrality = 'degree', return = "table")
```

The definition of the centrality metrics are as follows:

- **betweenness**: number of shortest paths going through a node.
- **closeness**: number of steps required to access every other node from a given node.
- **degree**: number of connections linked to a node.
- **eigenvector**: a measure of the influence a node has on a network.
- **pagerank**: calculates the PageRank for the specified vertices. Please refer to the igraph package documentation for the detailed technical definition.

### Communities

In the context of ONA, **community detection** refers to the process of identifying groups of nodes (in this case, individuals within an organization) that are more densely connected with each other than with the rest of the network. These groups, or communities, often reveal substructures within the organization, such as informal groups or teams that collaborate across department, job roles, or geographical boundaries.  

Community detection can provide valuable insights into the structure and dynamics of an organization. For example, it can help identify silos within the organization, where information or resources might not be flowing efficiently. It can also help identify key individuals who serve as bridges between different communities, who could be leveraged to improve communication and collaboration across the organization.

You can use the `community` and the `comm_args` argument in `network_p2p()` to perform community detection. Here is an example where we also specify the function to use the "ggraph" style: 
   
```{r message=FALSE, warning=FALSE}
network_p2p(
   data = p2p_data,
   community = "leiden",
   style = "ggraph",
   comm_args = list("resolution" = 0.1)
   )
``` 

Here, we are using the [Leiden algorithm](https://igraph.org/r/doc/cluster_leiden.html) to detect communities. The `resolution` parameter is a tuning parameter that controls the size of the communities. A higher value of `resolution` will result in smaller (and more) communities. `network_p2p()` operates on top of the community detection algorithms offered by [igraph](https://igraph.org/r/) package, with the following algorithms supported:

- **leiden** (`igraph::cluster_leiden()`): Leiden algorithm of Traag, van Eck, & Waltman
- **walktrap** (`igraph::cluster_walktrap()`): Walktrap community finding algorithm using short random walks, of Pons and Latapy
- **fast_greedy** (`igraph::cluster_fast_greedy()`): Fast greedy modularity optimization, of Clauset, Newman, & Moore
- **edge_betweenness** (`igraph::cluster_edge_betweenness()`): Edge betweenness community finding algorithm, of Girvan & Newman
- **infomap** (`igraph::cluster_infomap()`): Infomap community finding algorithm, of Rosvall & Bergstrom
- **label_prop** (`igraph::cluster_label_prop()`): Label propagation community finding algorithm, of Raghavan, Albert, & Kumara
- **leading_eigen** (`igraph::cluster_leading_eigen()`): Leading eigenvector community finding algorithm, of Newman
- **multilevel** (`igraph::cluster_multilevel()`): Multilevel community finding algorithm, of Blondel et al.
- **spinglass** (`igraph::cluster_spinglass()`): Spinglass community finding algorithm, of Reichardt & Bornholdt

Here is another example of how to use a different community detection algorithm: 
```{r message=FALSE, warning=FALSE}
network_p2p(
  data = p2p_data,
  style = "ggraph",
  community = "fluid_communities",
  comm_args = list("no.of.communities" = 5)
)
``` 

Please see the [original igraph documentation](https://igraph.org/r/doc/) for more details on the community detection parameters to pass into `comm_args`. 

You can also choose to visualize and explore the communities using other options available in `return`. For instance, you can return a sankey visual: 
```{r message=FALSE, warning=FALSE}
network_p2p(
  data = p2p_data,
  community = "leiden",
  return = "sankey",
  comm_args = list("resolution" = 0.1)
)
```

Or a table of the communities, and how they group together with the HR attribute specified in `hrvar`:
```{r message=FALSE, warning=FALSE}
network_p2p(
  data = p2p_data,
  community = "leiden",
  return = "table",
  comm_args = list("resolution" = 0.1)
)
```

## Analysis Approach

> The best practice towards network analysis is always to first establish and validate hypotheses with the data, and *then* visualize the result. The reason why one should not start an analysis by interpreting a network visual is that network visualizations are critically affected by inputs such as layout algorithms (of which many are non-deterministic), inputs to those layout algorithms, and visual elements such as the colours and transparencies of the nodes and edges. 

The above point on best practice is particularly pertinent for person-to-person analyses, which typically involve a large number of nodes and misleading conclusions can be drawn if one is not fully aware of the input parameters that have been used in generating the visual. In this case, the recommend interpretative approach would be to: 

1. Run the summary tables or graph statistics by returning `'table'` or `'network'`, making use of graph analysis functions that are compatible with an `'igraph'` object. 

2. Develop a series of hypotheses based on your knowledge of the organization and the data you are seeing. For instance, the hypothesis could be that Team A and Team B have very few ties with each other, but Region C maintain very strong ties with both of those regions, making it potentially a less optimal solution to allocate Team A and B together on ‘office day’ rotations. The same principle applies when generating communities. 

3. Validate hypothesis with the data, using multiple permutations. For instance, the selected date range and population could lead to a different outcome in the data. 

   Iterate with input parameters to create the visual that is most appropriate for communicating your analysis.

## End Note & Reference

This article is a first introduction to the topic of Organizational Network Analysis but there is more to come. At this point you should know the basic building blocks of network analysis and how to run the relevant functions in the **vivainsights** R library. Here are some more references on how to use the **vivainsights** R package and the topic of networks:

- [Overview of functions of the vivainsights R package](https://microsoft.github.io/vivainsights/reference/index.html)
- [Handbook of Graphs and Networks in People Analytics](https://ona-book.org/index.html)