---
title: "Radar Charts with create_radar()"
author: "vivainsights"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Radar Charts with create_radar()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
```

```{r load, message = FALSE}
library(vivainsights)
library(dplyr)
```

## Overview

`create_radar()` produces a multi-group radar (spider) chart that compares several metrics simultaneously across different segments of your workforce. The key steps inside the function are:

1. **Person-level aggregation** — averages (or medians) each metric per person per group.
2. **Group-level aggregation** — summarises person-level values to one row per group.
3. **Privacy filtering** — drops groups below `mingroup` unique persons.
4. **Optional indexing** — rescales values so groups are easy to compare.

---

## Basic usage

Pass a data frame, a character vector of metric columns, and a grouping column. By default the chart indexes each metric so that the overall population mean equals 100.

```{r basic-plot}
create_radar(
  data    = pq_data,
  metrics = c("Collaboration_hours", "Email_hours", "Meeting_hours"),
  hrvar   = "Organization",
  mingroup = 5
)
```

Each polygon represents one group. A value of 100 on any axis means that group sits exactly at the population average for that metric; values above 100 indicate above-average activity and values below 100 indicate below-average activity.

### Returning the underlying table

Set `return = "table"` to get the indexed values as a data frame instead:

```{r basic-table}
radar_tbl <- create_radar(
  data    = pq_data,
  metrics = c("Collaboration_hours", "Email_hours", "Meeting_hours"),
  hrvar   = "Organization",
  mingroup = 5,
  return  = "table"
)

radar_tbl
```

The `n` column records the number of unique persons in each group — useful for checking which groups are close to the privacy threshold.

---

## Indexing modes

The `index_mode` parameter controls how metric values are rescaled before plotting.

### `"total"` (default)

Each metric is divided by the overall population mean and multiplied by 100. A group at 100 matches the average.

```{r index-total}
create_radar(
  data       = pq_data,
  metrics    = c("Collaboration_hours", "Email_hours",
                 "Meeting_hours", "After_hours_collaboration_hours"),
  hrvar      = "Organization",
  mingroup   = 5,
  index_mode = "total"
)
```

### `"ref_group"` — benchmark against one group

Set `index_ref_group` to the name of the group that should equal 100 on every axis. All other groups are expressed relative to it.

```{r index-ref-group}
create_radar(
  data            = pq_data,
  metrics         = c("Collaboration_hours", "Email_hours", "Meeting_hours"),
  hrvar           = "Organization",
  mingroup        = 5,
  index_mode      = "ref_group",
  index_ref_group = "Finance"
)
```

### `"minmax"` — relative spread within observed groups

Each metric is scaled so the lowest group maps to 0 and the highest to 100. This maximises visual contrast and is useful when absolute levels matter less than which group is highest or lowest.

```{r index-minmax}
create_radar(
  data       = pq_data,
  metrics    = c("Collaboration_hours", "Email_hours", "Meeting_hours"),
  hrvar      = "Organization",
  mingroup   = 5,
  index_mode = "minmax"
)
```

### `"none"` — raw group values

No rescaling is applied. The axes carry the original units, which makes cross-metric comparison harder but preserves the absolute magnitude.

```{r index-none}
create_radar(
  data       = pq_data,
  metrics    = c("Collaboration_hours", "Email_hours", "Meeting_hours"),
  hrvar      = "Organization",
  mingroup   = 5,
  index_mode = "none",
  return     = "table"
)
```

---

## Aggregation method

The default two-step aggregation uses the **mean** at both the person level and the group level. Switch to `agg = "median"` for robustness against outliers.

```{r agg-median}
create_radar(
  data     = pq_data,
  metrics  = c("Collaboration_hours", "Email_hours",
               "Meeting_hours", "After_hours_collaboration_hours"),
  hrvar    = "Organization",
  mingroup = 5,
  agg      = "median"
)
```

---

## Grouping by a different HR variable

Any character column can serve as the grouping variable. Here we compare collaboration profiles by `LevelDesignation`:

```{r hrvar-level}
create_radar(
  data     = pq_data,
  metrics  = c("Collaboration_hours", "Email_hours",
               "Meeting_hours", "Internal_network_size"),
  hrvar    = "LevelDesignation",
  mingroup = 5
)
```

---

## Privacy filtering with `mingroup`

Groups with fewer than `mingroup` unique persons are silently dropped before plotting. Raise the threshold to be more conservative:

```{r mingroup}
# Only show groups with 20 or more unique persons
create_radar(
  data     = pq_data,
  metrics  = c("Collaboration_hours", "Email_hours", "Meeting_hours"),
  hrvar    = "Organization",
  mingroup = 20,
  return   = "table"
)
```

---

## Handling missing values

By default (`na.rm = FALSE`) the function retains rows with `NA` metric values and lets the aggregation handle them. Set `na.rm = TRUE` to drop any row containing an `NA` in any of the requested metrics before aggregation — this is equivalent to a complete-cases analysis.

```{r na-rm}
create_radar(
  data     = pq_data,
  metrics  = c("Collaboration_hours", "Email_hours"),
  hrvar    = "Organization",
  mingroup = 5,
  na.rm    = TRUE,
  return   = "table"
)
```

---

## Using the lower-level helpers directly

`create_radar()` is built from two exported helpers that you can call independently:

- **`create_radar_calc()`** — returns a list with `$table` (the indexed group-level data frame) and `$ref` (the reference values used for indexing).
- **`create_radar_viz()`** — accepts the `$table` output and returns a `ggplot` object.

This is useful when you want to post-process the table before plotting, or when you want to apply a custom theme on top of the default:

```{r low-level}
library(ggplot2)

calc <- create_radar_calc(
  data       = pq_data,
  metrics    = c("Collaboration_hours", "Email_hours", "Meeting_hours"),
  hrvar      = "Organization",
  mingroup   = 5,
  index_mode = "total"
)

# Inspect the reference values used for indexing
calc$ref

# Render with an extra annotation layer
create_radar_viz(
  data    = calc$table,
  metrics = c("Collaboration_hours", "Email_hours", "Meeting_hours"),
  hrvar   = "Organization"
) +
  labs(caption = "Index: 100 = population average")
```
