---
title: "Survival Curves with create_survival()"
author: "vivainsights"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Survival Curves with create_survival()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 8,
  fig.height = 5
)
```

```{r load, message = FALSE}
library(vivainsights)
library(dplyr)
```

## Overview

`create_survival()` produces a **Kaplan–Meier survival curve** — a non-parametric estimate of how quickly a specific event occurs across a population. In a workforce analytics context typical events include:

- First week of after-hours collaboration
- First week a collaboration or network metric crosses a threshold
- First observed use of a new tool

The function expects **one row per person** with a pre-computed `time` (weeks until event or end of observation) and `event` (1 = event occurred, 0 = censored) column. Use `create_survival_prep()` to derive these from a Standard Person Query panel dataset.

### Reframing "survival" for workforce contexts

In classical survival analysis the *event* is typically something negative — death, equipment failure — and the y-axis probability represents "still alive / still working". In workforce analytics the event is usually **a positive milestone**: first adoption of a tool, first week as a power user, first time a network or collaboration metric crosses a meaningful threshold.

The terminology inverts: "surviving" means *not yet having reached the milestone*, and the event ("death") means *success — the person converted*. It can therefore be more intuitive to read the chart as a **time-to-adoption curve**, a **conversion curve**, or a **graduation curve**:

- **A curve that drops steeply early** → most people reached the milestone quickly.
- **A curve that stays high** → many people had not yet converted by the end of the observation window.
- **The y-axis** → the share of people who have *not yet* experienced the milestone.

The curve answers: *"By week N, what fraction of the population had not yet crossed the threshold?"*

---

## Step 1 — Prepare person-level survival data

`pq_data` is a panel dataset with one row per person per week. `create_survival_prep()` collapses it to one row per person, recording:

- **`time`**: the week number at which the event first occurred, or the total number of observed weeks if the event never occurred (censored).
- **`event`**: 1 if the condition was met in at least one week, 0 otherwise.

Here we define the event as a person's **first week with any after-hours collaboration** (`After_hours_collaboration_hours > 0`):

```{r prep}
surv_data <- create_survival_prep(
  data            = pq_data,
  metric          = "After_hours_collaboration_hours",
  event_condition = function(x) x > 0,
  hrvar           = "Organization"
)

glimpse(surv_data)
```

```{r prep-summary}
# Event rate and time distribution
cat("Total persons:  ", nrow(surv_data), "\n")
cat("Event rate:     ", round(mean(surv_data$event) * 100, 1), "%\n")
cat("Weeks observed: ", range(surv_data$time), "\n")

surv_data %>%
  count(event, name = "n") %>%
  mutate(pct = round(100 * n / sum(n), 1),
         label = ifelse(event == 1, "Had event", "Censored"))
```

The persons who never had any after-hours work during the observation window appear as censored (`event = 0`). The survival curve accounts for this: they contribute information up to their last observed week.

---

## Step 2 — Plot the Kaplan–Meier curve

Pass the person-level data to `create_survival()`, specifying the time and event columns and the grouping variable:

```{r basic-plot}
create_survival(
  data      = surv_data,
  time_col  = "time",
  event_col = "event",
  hrvar     = "Organization",
  mingroup  = 5
)
```

**Reading the chart:**

- **Y-axis** — probability of *not yet* having reached the milestone (the "survival" probability, or equivalently, the share yet to convert).
- **X-axis** — week number (time since the start of observation).
- **Each step down** marks one or more conversions (events) in that group at that week.
- **Curves that drop quickly** indicate groups where most people reached the milestone early (fast adoption).
- **Curves that stay high** indicate groups where many people had not yet converted by the end of the window (slow or incomplete adoption).

---

## Overall curve (no grouping)

Set `hrvar = NULL` to estimate a single curve across the whole population:

```{r overall}
create_survival(
  data      = surv_data,
  time_col  = "time",
  event_col = "event",
  hrvar     = NULL,
  mingroup  = 5
)
```

---

## Returning the survival table

Set `return = "table"` to get the underlying long-format data frame. Each row represents one event time within one group:

```{r table}
surv_tbl <- create_survival(
  data      = surv_data,
  time_col  = "time",
  event_col = "event",
  hrvar     = "Organization",
  mingroup  = 5,
  return    = "table"
)

head(surv_tbl, 12)
```

Columns:

| Column | Description |
|--------|-------------|
| `Organization` | Group identifier |
| `n` | Total persons in group (after privacy filtering) |
| `time` | Week number |
| `survival` | Estimated survival probability at that time |
| `at_risk` | Persons still in the risk set (event not yet occurred) |
| `events` | Events occurring at this time point |

### Extracting median event times

The **median survival time** is the week at which 50 % of the group has experienced the event:

```{r median-time}
surv_tbl %>%
  group_by(Organization) %>%
  filter(survival <= 0.5) %>%
  slice(1) %>%                         # first time survival crosses 0.5
  ungroup() %>%
  select(Organization, n, time, survival) %>%
  arrange(time)
```

Groups with a smaller `time` value here reach the event threshold faster on average.

---

## Grouping by a different HR variable

Any character column can be used as the grouping variable. Here we compare after-hours adoption by `LevelDesignation`:

```{r hrvar-level}
surv_level <- create_survival_prep(
  data            = pq_data,
  metric          = "After_hours_collaboration_hours",
  event_condition = function(x) x > 0,
  hrvar           = "LevelDesignation"
)

create_survival(
  data      = surv_level,
  time_col  = "time",
  event_col = "event",
  hrvar     = "LevelDesignation",
  mingroup  = 5
)
```

---

## Changing the event definition

The `event_condition` argument in `create_survival_prep()` accepts any function that returns a logical vector. This makes it easy to explore different thresholds without modifying your data.

### Higher after-hours threshold

```{r event-threshold}
# Event: first week with more than 2 hours of after-hours collaboration
surv_high <- create_survival_prep(
  data            = pq_data,
  metric          = "After_hours_collaboration_hours",
  event_condition = function(x) x > 2,
  hrvar           = "Organization"
)

cat("Event rate (> 2 h after-hours):", round(mean(surv_high$event) * 100, 1), "%\n")

create_survival(
  data      = surv_high,
  time_col  = "time",
  event_col = "event",
  hrvar     = "Organization",
  mingroup  = 5
)
```

### Network growth milestone

```{r event-network}
# Event: first week where internal network size exceeds 10 contacts
surv_net <- create_survival_prep(
  data            = pq_data,
  metric          = "Internal_network_size",
  event_condition = function(x) x > 10,
  hrvar           = "Organization"
)

cat("Event rate (network > 10):", round(mean(surv_net$event) * 100, 1), "%\n")

create_survival(
  data      = surv_net,
  time_col  = "time",
  event_col = "event",
  hrvar     = "Organization",
  mingroup  = 5
)
```

---

## Privacy filtering

Groups below `mingroup` unique persons are removed before the curve is estimated. Increase the threshold to be more conservative:

```{r mingroup}
surv_strict <- create_survival(
  data      = surv_data,
  time_col  = "time",
  event_col = "event",
  hrvar     = "Organization",
  mingroup  = 20,
  return    = "table"
)

# Which groups remained after stricter filtering?
surv_strict %>%
  distinct(Organization, n) %>%
  arrange(desc(n))
```

---

## Handling missing values

`na.rm = TRUE` (the default) drops rows where `time_col` or `event_col` are `NA` before the curve is estimated. Set `na.rm = FALSE` to keep them; any remaining `NA` values will be silently excluded during coercion and a warning will be issued.

---

## Using the lower-level helpers directly

Like `create_radar()`, the survival family exposes its building blocks as exported functions:

- **`create_survival_calc()`** — accepts person-level data and returns a list with `$table` (the long survival table) and `$counts` (group size table).
- **`create_survival_viz()`** — accepts the `$table` output and returns a `ggplot` object.

```{r low-level}
library(ggplot2)

calc <- create_survival_calc(
  data      = surv_data,
  time_col  = "time",
  event_col = "event",
  hrvar     = "Organization",
  mingroup  = 5
)

# Inspect group sizes
calc$counts

# Render with extra annotation
create_survival_viz(
  data     = calc$table,
  hrvar    = "Organization",
  title    = "Time to first after-hours week",
  subtitle = "Kaplan\u2013Meier estimate by organisation"
) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  annotate("text", x = 0, y = 0.52, label = "50 % threshold",
           hjust = 0, size = 3, colour = "grey50")
```

The dashed line at 0.5 makes it easy to read off the median time-to-event for each group.
